---
title: "Gellman and Hill 13.2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

_Models for adjusting individual ratings:  a committee of 10 persons is evaluating 100 job applicants.  Each person on the committee reads 30 applications (structured so that each application is read by 3 people) and gives a numerical rating between 1 and 10._

_(a) It would be nautal to rate the applications based on their combined scores; however, there is aworry that different raters use different standards (severity) and we would like to correct for this.  Set up a model for the ratings (with parameters for the applicants and the raters)._

### Problem setup 
theta[i] -- Ability of Applicant i

eta[j] -- Severity of Rater j

epsilon[i,j] -- measurement error

Y[i,j] = theta[i]+eta[j] +epsilon[i,j]

Score ~ (1 | applicant) + (1 | rater)

### Design Matrix

There are `r choose(10,3)` ways of assigning 3 raters, so more patterns than we have rows.

```{r}
I <- 100L # applicants
J <- 10L  # raters
W <- 3L   # ratings per applicant
assignment <- matrix(0L,I,J)
for (i in 1L:I) {
  workload <- colSums(assignment)
  available <- which (workload < W*J)
  if (i > 75L)
    cat("Round ",i,": available = ",
        paste(available,collapse=", "),"\n")
  while (length(available) < W) {
    slacker <- which.min(workload)
    pswaps <- which(!assignment[1L:(i-1L),slacker])
    swaprow <- sample(pswaps,1L)
    swapcol <- sample(which(as.logical(assignment[swaprow,])))
    assignment[swaprow,swapcol] <- 0L
    assignment[swaprow,slacker] <- 1L
    workload <- colSums(assignment)
    available <- which(workload < W*J)
    cat("Round ",i,"x: availble=",paste(available,collapse=", "),
        "\n")
  } 
  assignment[i,sample(available,W)] <- 1L
}
colSums(assignment)
head(assignment)
```


